---
title: "R Notebook"
output: html_document
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.0'
      jupytext_version: 0.8.6
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
library(conflicted)
library(edgeR)
library(dplyr)
library(readr)
library(Matrix)
options(repr.matrix.max.cols=50, repr.matrix.max.rows=6)

```


```{r}
# r_mat = read_delim("../../../results/downstream_analysis/edger/r_mat.csv", delim=" ", col_names=FALSE)
r_mat = readMM("../../../results/downstream_analysis/edger/r_mat.mtx")
r_obs = read_csv("../../../results/downstream_analysis/edger/r_obs.csv", guess_max=100000)
r_genes = read_csv("../../../results/downstream_analysis/edger/r_genes.csv")
```

```{r}
r_obs2 = r_obs %>% 
  mutate(dataset = as.factor(dataset),
         clusters = as.factor(paste0("cluster_", clusters)),
         tumor_type = if_else(tumor_type == "LUAD", "NSCLC", tumor_type))

```


```{r}
design = model.matrix(~0+clusters + dataset + n_genes + percent_mito, data=r_obs2)
```

```{r}
contrasts = makeContrasts(
      cluster0 = clusterscluster_0 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster1 = clusterscluster_1 - (clusterscluster_0+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster2 = clusterscluster_2 - (clusterscluster_1+clusterscluster_0+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster3 = clusterscluster_3 - (clusterscluster_1+clusterscluster_2+clusterscluster_0+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster4 = clusterscluster_4 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_0+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster5 = clusterscluster_5 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_0+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster6 = clusterscluster_6 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_0+
                                     clusterscluster_7+clusterscluster_8)/8
    , cluster7 = clusterscluster_7 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_0+clusterscluster_8)/8
    , cluster8 = clusterscluster_8 - (clusterscluster_1+clusterscluster_2+clusterscluster_3+
                                     clusterscluster_4+clusterscluster_5+clusterscluster_6+
                                     clusterscluster_7+clusterscluster_0)/8 
    , levels=colnames(design)
)
```

```{r}
dge = DGEList(counts=r_mat, samples=r_obs2, genes=r_genes)
```

```{r}
dge <- calcNormFactors(dge)
```

```{r}
dge <- estimateDisp(dge, design = design)
```

```{r}
fit <- glmQLFit(dge, design = design)
```

```{r}
qlfs = sapply(colnames(contrasts), function(cluster) {
    message(paste0("working on ", cluster))
    glmQLFTest(fit, contrast=contrasts[,cluster])
}, USE.NAMES=TRUE, simplify=FALSE)
```

```{r}
tts = sapply(qlfs, function(qlf) {
    topTags(qlf, n=Inf, adjust.method="BH")
}, USE.NAMES=TRUE, simplify=FALSE)
```

```{r}
all_results = bind_rows(lapply(names(tts), function(cluster) {
    tts[[cluster]]$table %>% mutate(cluster=cluster)
})) %>% as_tibble()
```

```{r}
save(design, dge, fit, qlfs, tts, all_results, file="../../../results/downstream_analysis/edger/results.rda", compress = FALSE)
```

```{r}
write_tsv(all_results, "../../../results/downstream_analysis/edger/de_genes.tsv")
```

```


## repeat the analysis for each tumor type individually

```{r}
tumor_types = r_obs2 %>% pull(tumor_type) %>% unique()
```

```{r}
sapply(tumor_types, function(tumor_type) {
    
    # prepare data and bulid design matrix
    message(paste0("# Working on ", tumor_type))
    tmp_obs = r_obs2 %>% filter(tumor_type == !!tumor_type) %>%
        mutate(clusters=droplevels(clusters)) %>%
        mutate(dataset=droplevels(dataset))
    if(tumor_type == "LIHC") {
        # single dataset, all percent_mito == 0
        design = model.matrix(~0+clusters + n_genes, data=tmp_obs)
    } else {
        design = model.matrix(~0+clusters + dataset + n_genes + percent_mito, data=tmp_obs)
    }
    
    # build contrasts as diagonal matrix
    contrasts = sapply(1:nlevels(tmp_obs$clusters), function(i) {
        n = nlevels(tmp_obs$clusters)
        v = rep(0, length(colnames(design)))
        names(v) = colnames(design)
        v[1:n] = -1/(n-1)
        v[i] = 1
        v
    })
    colnames(contrasts) = levels(tmp_obs$clusters)
    
    # run DE analysis. 
    dge = DGEList(counts=r_mat[,r_obs2$tumor_type == tumor_type], 
                  samples=tmp_obs, genes=r_genes)
    dge <- calcNormFactors(dge)
    message("estimateDisp")
    dge <- estimateDisp(dge, design = design)
    message("glmQLFit")
    fit <- glmQLFit(dge, design = design)
    qlfs = sapply(colnames(contrasts), function(cluster) {
        message(paste0("working on QLFTest for ", cluster))
        glmQLFTest(fit, contrast=contrasts[,cluster])
    }, USE.NAMES=TRUE, simplify=FALSE)
    tts = sapply(qlfs, function(qlf) {
        topTags(qlf, n=Inf, adjust.method="BH")
    }, USE.NAMES=TRUE, simplify=FALSE)
    all_results = bind_rows(lapply(names(tts), function(cluster) {
        tts[[cluster]]$table %>% mutate(cluster=cluster)
    })) %>% as_tibble()
    
    # save results
    write_tsv(all_results, paste0("../../../results/downstream_analysis/edger/de_genes_", tumor_type, ".tsv"))
    save(design, dge, fit, qlfs, tts, all_results,
         file=paste0("../../../results/downstream_analysis/edger/results_", tumor_type, ".rda"), 
         compress = FALSE)
}, USE.NAMES=FALSE, simplify=TRUE)
```

```{r}

```
